D3692R1: How to Avoid OOTA Without Really Trying

### 2025-06-17 paper/ootavacuous.tex
ootavacuous: Section 7.6: Wordsmith away redundancy
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 7.6 further wordsmithing
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add offlist discussions to history
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Comment out rhetorical question for Sofia meeting
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add footnote for Section 7.4 litmus test
Also add the missing "repurpose" to the proposed standard addition.

Sign

\begin{itemize}
\item	D3692R1 adds contact information, defines ``preserved
	dependencies'' that are a subset of syntactic and a superset of
	semantic dependencies, and makes numerous changes in the name
	of wordsmithing and clarification.
\item	This paper is a condensation of P3064R2 (``How to Avoid OOTA
	Without Really
	Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
	which was criticized as being overly long.
	It also adds citations and clarifications.
\end{itemize}

@@

In short, this paper's contribution is to show how these
constraints allow real-world C++ implementations doing per-thread
analysis to avoid OOTA behavior easily,
%and not with minimal cost~\cite{PiezhaoOu2018OOTApapl},
in many cases with no cost at
all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
\emph{@@@ Are these the right qualifiers?}

@@

Although this result is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us because it is
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.

@@

Sensible though this recommendation might be, it can unnecessarily raise
anxiety levels of C++ implementers.
After all, what exactly do they need to do in order to follow that
recommendation?
One way to reduce their anxiety levels would be to add something like
the following sentence to this recommendation:
\begin{quote}
	Compiler-based implementations whose binaries run on conventional
	hardware are guaranteed not to compute out-of-thin-air values in
	programs that are free of undefined behavior,
	as long as they restrict themselves to thread-at-a-time analysis
	and do not invent or duplicate atomic accesses.
\end{quote}

>>>>

\begin{itemize}
\item	D3692R1 adds contact information, defines ``preserved
	dependencies'' that are a subset of syntactic and a superset of
	semantic dependencies, and makes numerous changes in the name
	of wordsmithing and clarification.
\item	This paper is a condensation of P3064R2 (``How to Avoid OOTA
	Without Really
	Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
	which was criticized as being overly long.
	Voluminous off-list discussions are reflected.
	It also adds citations and clarifications.
\end{itemize}

@@

In short, this paper's contribution is to show how these
constraints allow real-world C++ implementations doing per-thread
analysis to avoid OOTA behavior easily,
%and not with minimal cost~\cite{PiezhaoOu2018OOTApapl},
in many cases with no cost at
all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
% \emph{@@@ Are these the right qualifiers?}

@@

Although this result is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us because it is
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.\footnote{
	At the 2025 C++ Standards Committee meeting in Sofia, it was
	pointed out that the added ``\co{if (r1 != r3)}'' can be argued
	to be repurposing the first and third relaxed atomic load from
	\co{x}.
	At first glance, this is a very reasonable argument that would
	unambigously rule out this optimization.}

@@

Sensible though this recommendation might be, it can unnecessarily raise
anxiety levels of C++ implementers.
After all, what exactly do they need to do in order to follow that
recommendation?
We therefore advocate adding the following non-normative sentence to this
recommendation:
\begin{quote}
	Compiler-based implementations whose binaries run on conventional
	hardware are guaranteed not to compute out-of-thin-air values in
	programs that are free of undefined behavior,
	as long as they restrict themselves to thread-at-a-time analysis
	and do not invent, duplicate, or repurpose atomic accesses.
\end{quote}
